---
title: "Point patterns - estimation and modelling disease data: Portfolio"
author: "YOUR NAME HERE"
date: "Week 12"
output: html_document
---

## Introduction

In this worksheet we will look at how we take measurement information from a set of points and extend or interpolate that measurement across a spatial region.

The key idea is that measurements over a spatial region tend to be correlated - points that are close to each other typically give similar measurements because they are close.

So when we model the measurement spatially, we need to account for that correlation.

In this portfolio we will look at simple kriging and universal kriging.

**You should work through the worksheet_week11.Rmd worksheet before this one. You should then work through this worksheet and fill in your answers, Knit, and hand the .html into the dropbox on stream.**

## Packages

Today we'll be using the `tidyverse` for ggplot charts, and `sf` and `gstat` for our spatial analyses. You may need to install the `gstat` package. Let's load them:

```{r, message=FALSE}
library(tidyverse)
library(sf)
library(gstat)
```

## Data

The data for this portfolio is the pricing of 2 bedroom homes or apartments available for
rent from AirBNB in and around the Sydney central business district.

The data were obtained from http://insideairbnb.com/get-the-data and have been processed so as
to make an interesting spatial analysis case study.

We have coordinates of each AirBnB location along with their price, plus the distance to the harbour or coastline (i.e. distance to water).

In addition, we have the sydney harbour and coastline to provide context for our map, and a grid of locations across the region for us to interpolate to.

Let's load the data off the web and setup the CRS to a planar system appropriate for the area we're modelling (GDA2020 / MGA zone 56), and take a quick look at the sydney data:

```{r, message=FALSE}
sydney <- read_csv("https://www.massey.ac.nz/~jcmarsha/233214/data/airbnb/sydney.csv") |>
  st_as_sf(coords = c('X', 'Y'), crs=7856)

sydney_grid <- read_csv("https://www.massey.ac.nz/~jcmarsha/233214/data/airbnb/sydney_grid.csv") |>
  st_as_sf(coords = c('X', 'Y'), crs=7856)

coast <- read_sf("https://www.massey.ac.nz/~jcmarsha/233214/data/airbnb/sydney_coast.geojson")

sydney
```
 
You should see from the data that for each location we have an identifier, the name of the property, the price, and the distance to the coast.

The goal will be to find a geospatial statistical model to estimate the price of 2 bedroom AirBnB properties at other locations across the area, noting that:

1. There is very likely spatial auto-correlation: if a property has a higher price then it is likely near other properties with high prices (e.g. there is a benefit of location to what people will pay for a property).

2. There is likely a relationship between the price per night and how close the property is to the harbour or coastline. People are likely to pay more for a water view.

## Data exploration

We'll start by plotting the location data using ggplot:

```{r}
ggplot(sydney |> arrange(price)) +
  geom_sf(aes(col = price), alpha=0.5) +
  geom_sf(data=coast) +
  scale_colour_continuous(trans=scales::log10_trans())
```

### Q1: Based on this plot, do you see evidence of spatial autocorrelation in price? Explain your answer.

WRITE YOUR ANSWER HERE

Let's check whether there might be a relationship between distance to the coast and price, by ignoring the spatial component and just plotting price against distance to coast:

```{r}
ggplot(sydney) +
  geom_point(aes(x=dist_to_coast, y=price))
```

We note the relationship appears to be decreasing, but it is non-linear. In statistics generally if we can transform something to be linear it allows us to use the linear model - let's try a log/log plot instead:

```{r}
ggplot(sydney) +
  geom_point(aes(x=log(dist_to_coast), y=log(price)))
```

### Q2: In the code block below, fit a linear model for log(price) in terms of log(dist_to_coast) and produce a summary. Is the relationship useful? Explain your answer.

```{r}
```

WRITE YOUR CODE IN THE ABOVE BLOCK, AND YOUR ANSWER HERE

We'll now use simple kriging to interpolate the price over the gridded area. We start by doing a variogram and fitting a spherical variogram model to it:

```{r}
const.vgm <- variogram(log(price) ~ 1, data=sydney)
const.mod <- fit.variogram(const.vgm, model=vgm(0.5, "Sph", 5000, 0.1))
plot(const.vgm, const.mod)
```

Next, we perform kriging to the gridded surface and plot:

```{r}
const.krig <- krige(log(price) ~ 1, locations=sydney,
                    newdata=sydney_grid, model=const.mod)

ggplot(const.krig) +
  geom_sf(aes(col=exp(var1.pred))) +
  geom_sf(data=coast) +
  scale_color_viridis_c(trans = 'log10', limits=c(80, 3000)) +
  labs(col = 'Price per night')
```

Next, we'll try universal kriging to incorporate the information about distance to coast. Firstly the variogram, again using a spherical model:

```{r}
covar.vgm <- variogram(log(price) ~ log(dist_to_coast), data=sydney)
covar.mod <- fit.variogram(covar.vgm, model=vgm(0.5, "Sph", 5000, 0.1))
plot(covar.vgm, covar.mod)
```

### Q3: In the code block below, write the code to perform universal kriging of log(price) using log(dist_to_coast) and produce a plot of the resulting predicted prices.

```{r}
```

Suppose a new 2 bedroom apartment has been built and the owner is interested in using it for an AirBnB. The location is given below, where we also compute the distance to the coast.

```{r}
new_apartment <- data.frame(X = 335000, Y = 6251500) |>
  st_as_sf(coords = c("X", "Y"), crs=7856) |>
  mutate(dist_to_coast = st_distance(geometry, coast) |> as.numeric())
```

### Q4: Using the universal kriging model, What would you expect the price for this property to be?

```{r}
```

WRITE THE CODE IN THE ABOVE CHUNK AND SUMMARISE YOUR ANSWER HERE.
