---
title: "Simple Kriging for spatial interpolation"
date: "Week 12"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(sp)
library(gstat)
library(sf)
library(patchwork)

data(meuse.riv)
river <- data.frame(meuse.riv) %>% set_names(nm=c('x', 'y')) %>%
  filter(y < 334500 & y > 329500) %>%
  as.matrix() %>% st_linestring()

knitr::opts_chunk$set(echo = FALSE, message=FALSE, fig.align='center')
theme_set(theme_void())
```

## Introduction

Spatial interpolation is the process of taking measurement information from a set of points and extending or interpolating that measurement across a spatial region.

The key idea is that measurements over a spatial region tend to be correlated - points that are close to each other typically give similar measurements because they are close.

So when we model the measurement spatially, we need to account for that correlation.

There are a wide range of methods for spatial interpolation, and we'll cover [inverse distance weighting](week12_idw.html), simple kriging and [universal kriging](week12_universal_kriging.html).

As an example dataset, we'll be using a famous dataset of heavy metal concentrations in topsoil in a flood plain of the river Meuse near the village of Stein in the Netherlands. It is available within the `sp` package and you can find more details about it using `?meuse`. We'll be concentrating on the metal `zinc`, and look at it on a log scale:

```{r}
data(meuse)
meuse.sf <- st_as_sf(meuse, coords = c('x', 'y'))

ggplot(meuse.sf) +
  geom_sf(data=river) +
  geom_sf(mapping=aes(fill=log(zinc)), shape=21) +
  scale_fill_distiller(palette="YlGnBu", limits=c(4.4, 7.6))
```

The goal will be to interpolate this measure so we can estimate the concentration of zinc across the entire flood plain. To do this we'll make use of a grid for interpolation:

```{r}
data(meuse.grid)
meuse.grid.sf <- st_as_sf(meuse.grid, coords = c('x', 'y'))

ggplot(meuse.grid.sf) +
  geom_sf(data=river) +
  geom_sf(size=1, shape=21)
```

## Simple Kriging

An alternate, and perhaps more statistical way of performing interpolation is to use a statistical model to define how pairs of measurements are expected to be related, given how far they are apart. We do this by creating a **variogram** which takes half the squared difference in the measure $z$ between pairs of observations $i$, $j$:
$$
\gamma_{ij} = \frac{1}{2}(z_i - z_j)^2
$$
and plots this against the distance $d_{ij}$ between observations $i$ and $j$, for all pairs of observations up to some given distance (e.g. up to say 1/3 of the area covered by the points).

```{r}
d <- st_distance(meuse.sf)
gamma <- dist(log(meuse.sf$zinc)) %>% as.matrix()
gamma_vs_d <- data.frame(gamma = as.numeric(gamma)/2, dist = as.numeric(d)) %>%
  filter(dist < 1500)
ggplot(gamma_vs_d) +
  geom_point(aes(x=dist, y=gamma), alpha=0.3) +
  theme_minimal()
```

This is a bit overwhelming to interpret, so we typically divide the distance into bands and then estimate the average $\gamma$ within each band:

```{r}
bands <- gamma_vs_d %>% mutate(band = cut(dist, breaks=15)) %>%
  group_by(band) %>% summarise(dist=mean(dist), gamma=mean(gamma))

ggplot(gamma_vs_d) +
  geom_point(aes(x=dist, y=gamma), alpha=0.1) +
  geom_point(data=bands, mapping=aes(x=dist, y=gamma), size=4, col='dark red') +
  theme_minimal()
```

This can be done using the `variogram` function:

```{r, echo=TRUE}
const.var <- variogram(log(zinc) ~ 1, data=meuse.sf)
```

We then fit a **variogram model** to this. There's a whole heap of variogram models available:

```{r}
all <- show.vgms(plot=FALSE) %>% rename(gamma=semivariance)
ggplot(data=all) +
  geom_line(mapping=aes(x=distance, y=gamma)) +
  facet_wrap(vars(model)) +
  theme_minimal()
```

We usually pick one of these models, then control it's fit to the data by modifying three main parameters: the **nugget**, the **partial sill** and the **range**. The nugget is the y-intercept (the value of $\gamma$) when distance = 0), while the range and partial sill are the point (x and y respectively) where the variogram starts flattening out.

This fitting is done in the `fit.variogram` function. Here we use the spherical variogram model, and we give it a guess at the partial sill (0.6), the range (1000) and the nugget (0.1). It then optimises these parameters from there and gives:

```{r, echo=TRUE}
const.mod <- fit.variogram(const.var, model=vgm(0.6, "Sph", 1000, 0.1))
const.mod
```

We were pretty close with our guesses. We can plot the result on top of our data variogram:

```{r}
ggplot(data=variogramLine(const.mod, maxdist=max(const.var$dist)),
       mapping=aes(x=dist, y=gamma)) +
  geom_line() +
  geom_point(data=const.var) +
  theme_minimal()
```

Now that we have a model for how pairs of points will be correlated, we can now do the interpolation. Essentially we use the variogram model to provide localised weighting parameters.

Recall that with [inverse distance weighting](week12_idw.html) the interpolated value at an unsampled site was determined by the weighted average of neighbouring points where the weighting parameter (the power $d$ in the formula) was the same across the entire study extent. Kriging uses the variogram model to compute the weights of neighboring points based on the distribution of those values, so it allows the local pattern of points in the neighbourhood of the unsampled point to define the weights. The weights $\mathbf{w} = [w_1, w_2, \ldots w_n]$ are determined using
$$
\mathbf{w} = \Gamma^{-1} \mathbf{\gamma_x}
$$
where $\Gamma = \gamma(D)$ is a matrix produced by applying the variogram model $\gamma$ to the pair-wise distance matrix $D$ of the sampled locations, and $\mathbf{\gamma_x} = \gamma(\lVert\mathbf{x} - \mathbf{x}_i\rVert)$ is the vector produced by applying the variogram function to the vector of distances from each sampled location to the unsampled location we are interpolating to.

Once we have the weights $\mathbf{w}$, we then compute the interpolated measure $\hat{Z}$ using the same equation as was used for [IDW](week12_idw.html):
$$
\hat{Z} = \frac{\sum_{i=1}^n w_iz_i}{\sum_{i=1}^n w_i}
$$
We do this in R with the `krige` function:

```{r, echo=TRUE, eval=FALSE}
krig.const <- krige(log(zinc) ~ 1, meuse.sf, meuse.grid.sf, model=const.mod)
```

which gives us an `sf` object with the same columns (`var1.pred` and `var1.var`) as the `idw` function did. The difference is that `var1.var` is now useful as a measure of uncertainty. You should see low uncertainty where we have observations, and higher uncertainty where we don't.

```{r}
krig.const <- krige(log(zinc) ~ 1, meuse.sf, meuse.grid.sf, model=const.mod,
                    debug.level = 0)
g1 <- ggplot(krig.const) +
  geom_sf(data=river) +
  geom_sf(mapping=aes(col=var1.pred), size=0.7) +
  scale_color_distiller(palette="YlGnBu", limits=c(4.4, 7.6)) +
  labs(col="log(zinc)")
g2 <- ggplot(krig.const) +
  geom_sf(data=river) +
  geom_sf(mapping=aes(col=var1.var), size=0.7) +
  scale_color_distiller(palette="YlOrBr") +
  labs(col="uncertainty")
g1 + plot_spacer() + g2 + plot_layout(widths=c(1,0.1,1))
```
