---
title: "Spatial interpolation"
date: "Week 12"
output:
  html_document:
    code_download: true
---

## Introduction

In this worksheet we will look at how we take measurement information from a set of points and extend or interpolate that measurement across a spatial region.

The key idea is that measurements over a spatial region tend to be correlated - points that are close to each other typically give similar measurements because they are close.

So when we model the measurement spatially, we need to account for that correlation.

We'll look at three methods: Inverse distance weighting, simple kriging and universal kriging.

**You should work through this worksheet, but do not need to hand it in as part of your Portfolio.**

## Packages

Today we'll be using the `tidyverse` for ggplot charts, and `sf` and `gstat` for our spatial analyses. You may need to install the `gstat` package. Let's load them:

```{r, message=FALSE}
library(tidyverse)
library(sf)
library(gstat)
```

## Data

The data for this worksheet is on the mass of bats in North America. It was put together for this paper:

https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.7641

which concerns the disease White-nose syndrome (WNS) in little brown myotis (Myotis lucifugus) in North America.

The disease effects bats while they are hibernating over winter, essentially causing them to come out of torpor which requires a lot of energy. Thus, the bats that survive tend to be those that have the required body mass (fat) to survive this over the winter period.

As body mass is a key requirement, we were interested in estimating the likely body mass of bats across North America which would then feed into an energetics model alongside winter duration in order to estimate survival of bats.

The data are available at only 42 locations, so this is a bit of a stretch! Nonetheless, let's see what we can do. We start by reading in the data locations along with the north american boundary, setup the CRS and do a plot, choosing Lambert conformal conic projection: 

```{r, message=FALSE}
# read in the data
bats.csv = read_csv("https://www.massey.ac.nz/~jcmarsha/233214/data/bats/bats.csv")
NA_boundary <- read_sf("https://www.massey.ac.nz/~jcmarsha/233214/data/bats/NA_boundary.sqlite")
bats = st_as_sf(bats.csv, coords=c("long","lat"), crs=st_crs(NA_boundary))

# set up a Lambert conformal conic projection for lat/long
crs_for_map <- "+proj=lcc +lat_0=0 +lon_0=-95 +lat_1=49 +lat_2=77 +datum=WGS84 +units=m"

# plot our bat locations on the map
ggplot() +
  geom_sf(data=NA_boundary) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map)
```

We can see that they're largely scattered across the US and Canada. Our coordinates are latitude and longitude, so we note that this isn't ideal for doing spatial analyses as we'd generally prefer a planer coordinate system when using distances etc. However, the functions we'll be using today for inverse distance weighting and Kriging will correctly use 'great circle' distances.

## Covariate information

We have covariate information, including things like the number of freezing days, growing days etc available to us. These have been derived from the Adaptwest project:

https://adaptwest.databasin.org/pages/adaptwest-climatena/

We've downsampled them to make things a bit more manageable for today's workshop. The next code block reads them in, and adds in an additional geometry layer which is the centroid of each grid area. This will be useful for our spatial interpolation.

```{r}
NA_info <- read_sf("https://www.massey.ac.nz/~jcmarsha/233214/data/bats/NA_info.sqlite")
NA_points <- NA_info |>
  mutate(tile = GEOMETRY,
         GEOMETRY = st_centroid(tile))

NA_points
```

As you can see, we have data on elevation, northing (latitude), frost days, freezing days and growing days. All of these might be useful for estimating the mass of bats.

Let's plot one of these:

```{r}
ggplot() +
  geom_sf(data=NA_points, aes(geometry=tile, fill=growing_days), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  scale_fill_viridis_c() +
  coord_sf(crs = crs_for_map)
```

### Q1: Change the code above to plot some other variables, e.g. `frost_days`, `growing_days`, `elevation` and `northing`. Which ones do you think correlate with bat mass?

Northing, frost days, freezing days all correlate with bat mass, but elevation doesn't. Growing days also correlates negatively with bat mass.

## Inverse distance weighted interpolation

Before we get into spatial modelling for interpolation, let's start with a relatively simple method of spatial interpolation based on inverse distance. The idea is that we interpolate to a point by averaging across our observations, weighted by the distance from our point to each observation: Observations that are far away will be weighted less while observations that are close are weighted more.

We can use the `idw` function from the `gstat` package for this. We set the inverse distance power to be 2, which means we're using one over the distance squared as our weights.

NOTE: If you have the `spatstat` package loaded, this might not work as there may be a clash with `spatstat::idw`. You can change the `idw` below to `gstat::idw` as one way to fix this.

```{r}
mass.idw <- gstat::idw(mass ~ 1, locations=bats, newdata=NA_points,
                idp=2)
mass.idw
```

Notice that this gives us an `sf` object again with two columns `var1.pred` and `var1.var` (unused, so set to `NA`) along with the `geometry` column. We could plot these points as is:

```{r}
ggplot(mass.idw) +
  geom_sf(mapping=aes(col=var1.pred)) +
  scale_colour_viridis_c() +
  coord_sf(crs = crs_for_map)
```

but it will likely look a bit nicer if we join back to our `NA_points` dataset so we can use the `tile` geometry for plotting:

```{r}
# spatial join to the NA_points dataset
NA.idw.tile <- mass.idw |> st_join(NA_points)

# And plot
ggplot() +
  geom_sf(data=NA.idw.tile, aes(geometry=tile, fill=var1.pred), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_viridis_c(limits = c(5,13))
```

As might be expected, we have basically a smoothed map where the predicted mass converges on the mass at each location.

In the code block below, try taking the above code and adapting it for the case where the inverse distance power is 1 (i.e. set `idp=1` in the `idw` function)

```{r}
# Do the IDW interpolation
mass.idw.1 <- gstat::idw(mass ~ 1, locations=bats, newdata=NA_points,
                idp=1)

# spatial join to the NA_points dataset
NA.idw.tile.1 <- mass.idw.1 |> st_join(NA_points)

# And plot
ggplot() +
  geom_sf(data=NA.idw.tile.1, aes(geometry=tile, fill=var1.pred), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_viridis_c(limits = c(5,13))
```

### Q2: Which inverse distance power do you think works best here?

The power of 2 seems to work a bit better - it doesn't overly smooth like a power of 1 does, but still allows some smoothing. Increasing the power to a bit higher gives even more prominence to the local data. A power of 2 seems a happy median in this case.

## Simple Kriging

An alternative to inverse distance weighting is simple kriging. What this does is applies a **variogram** model to account for the fact that at smaller distances the measure of interest (bat mass) is likely to not vary much, while at larger distances we might expect that variation to increase.

Essentially we estimate the pair-wise variation in bat mass in different distance bands, and then fit a mathematical model to those estimates so that we can get an idea of the variation in between the bands. The first step is done with the `variogram` function in `gstat`. The `cutoff` parameter means we don't want to go over 5000km in the evaluation of the variogram.

```{r}
const.vgm <- variogram(mass ~ 1, data=bats, cutoff=5000)
plot(const.vgm)
```

We can see that the variance is low at smaller distances (less than 1000 km) and then increases and bounces about a bit. Note that because we don't have many points, this is a super noisy measure. So at best we're going to have a pretty lousy estimate of how the variation in bat mass changes with distance.

We can attempt to fit a mathematical model to these estimates using the `fit.variogram` function:

```{r}
const.mod <- fit.variogram(const.vgm, model=vgm(3, "Exp", 3000, 1))
const.mod
plot(const.vgm, const.mod)
```

As you can see, this is a pretty rubbish fit at larger distances. But, the idea is OK - we expect that the variation in mass should increase as locations are further apart.

There are lots of variogram models - this is just one. There's a couple of key parameters - the nugget and the sill. The nugget is basically the variation at distance zero, and the sill is where it starts flattening out. So you can see here that our nugget has been estimated at 1.26 at distance zero, and the sill at 2.76 when the range is 1193km. The exponential curve we've fit still keeps increasing past the sill, but the coordinates (2.76, 1193) are governing the shape.

Once we have an estimate of the variogram, we can then fit a model to the data and utilise the variogram to interpolate to other locations. This is done using `krige`, where we pass the new data we want to interpolate to and the model we want to use for the variogram:

```{r}
const.krig <- krige(mass ~ 1, locations=bats, newdata=NA_points, model=const.mod)
const.krig
```

This is an `sf` object containing the outcome variable for prediction `var1.pred` and it's uncertainty measure `var1.var`, along with the geometry. Let's again join it back to get our tile geometry again and plot:

```{r}
const.tile <- const.krig |> st_join(NA_points)
ggplot() +
  geom_sf(data=const.tile, aes(geometry=tile, fill=var1.pred), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_viridis_c(limits=c(5,13))
```

We can also get uncertainty out of this, by plotting the `var1.var` variable:

```{r}
ggplot() +
  geom_sf(data=const.tile, aes(geometry=tile, fill=var1.var), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_distiller(palette="YlOrBr")
```

## Universal Kriging with covariates

Now, we know we have some covariates available, and back near the start we plotted these and checked for correlation with mass. We could also do this non-spatially:

```{r}
ggplot(data=bats) + geom_point(aes(x=freeze_days, y=mass))
```

Do similar plots for the `frost_days` and `growing_days` variables.

```{r}
ggplot(data=bats) + geom_point(aes(x=growing_days, y=mass))
ggplot(data=bats) + geom_point(aes(x=frost_days, y=mass))
ggplot(data=bats) + geom_point(aes(x=northing, y=mass))
ggplot(data=bats) + geom_point(aes(x=elevation, y=mass))
```

### Q3: Which variable(s) do you think are most useful for predicting the mass of bats?

Probably the freezing days is most useful. This will be when the bats are hibernating anyway, so presumably bat mass will be correlated with this (bats in regions that have longer winters will be heavier as they have to be to survive).

We could test some using a linear model or regression:

```{r}
mod_freeze <- lm(mass ~ freeze_days, data=bats)
summary(mod_freeze)
```

From here we note that `freeze_days` is important (it's P-value of 0.00303 is low), and it explains around 20% of the variation in mass (from the Multiple R-squared).

Try linear models for `frost_days` and `growing_days` as well.

```{r}
mod_frost <- lm(mass ~ frost_days, data=bats)
summary(mod_frost)
```

```{r}
mod_grow <- lm(mass ~ growing_days, data=bats)
summary(mod_grow)
```

```{r}
mod_elev <- lm(mass ~ elevation, data=bats)
summary(mod_elev)
```

### Q4: Which one explains the most of the variation in mass (i.e. which has the highest Multiple R-squared?)

Freezing days seems to explain the most variation in bat mass (around 20%), followed by frost days (around 19%) and growing days (around 10%). Elevation doesn't explain much at all (0.5%), so won't be useful.

We can include more than just one thing at a time if we want:

```{r}
mod_more <- lm(mass ~ freeze_days + frost_days + northing, data=bats)
summary(mod_more)
```

By adding more things into the model we explain more variation in bat mass (now 28%). Note the significance of individual covariates should be interpreted as being over and above the other covariates. i.e. does freeze_days make a difference if we've already accounted for frost_days and northing? (Yes - a little, the p-value is 0.04). Similarly, does northing make a difference after accounting for frost_days and freeze_days? Not really - the p-value is 0.12. Because all these things are correlated, we'd expect this to happen - that doesn't mean we should drop them out, as the goal here is good prediction - i.e. we should count the R^2 value more than the individual P-values for covariates when determining the 'best' model to use.

Ok, now these models don't account for the spatial correlation, but kriging can do this for us. Let's rebuild our variogram and kriging model using the `freeze_days` variable. First off, the variogram:

```{r}
formula <- mass ~ freeze_days
covar.vgm <- variogram(formula, locations=bats, cutoff=5000)
covar.mod <- fit.variogram(covar.vgm, model=vgm(1, "Exp", 3000, 1))
plot(covar.vgm, covar.mod)
```

This fit might be a bit better than our last one. We're underestimating the variation at very large distance, but capturing it at smaller distances OK. Now the kriging:

```{r}
covar.krig <- krige(formula, locations=bats, newdata=NA_points, model=covar.mod)

covar.tile <- covar.krig |> st_join(NA_points)
ggplot() +
  geom_sf(data=covar.tile, aes(geometry=tile, fill=var1.pred), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_viridis_c(limits=c(5,13))
```

When we do this we're making a much stronger assumption about the underlying trend, so we need to be a little cautious when predicting places that are away from our points. We can see this when we plot the uncertainty:

```{r}
ggplot() +
  geom_sf(data=covar.tile, aes(geometry=tile, fill=var1.var), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_distiller(palette = "YlOrBr", limits=c(0.5,4))
```

Just for fun, let's try our more complicated model with frost_days, freeze_days and northing:

```{r}
formula_big <- mass ~ freeze_days + frost_days + northing
big.vgm <- variogram(formula_big, locations=bats, cutoff=5000)
big.mod <- fit.variogram(big.vgm, model=vgm(1, "Exp", 3000, 1))
plot(big.vgm, big.mod)
```

This fits OK. Now our kriging

```{r}
big.krig <- krige(formula_big, locations=bats, newdata=NA_points, model=big.mod)

big.tile <- big.krig |> st_join(NA_points)
ggplot() +
  geom_sf(data=big.tile, aes(geometry=tile, fill=var1.pred), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_viridis_c(limits=c(5,13))
```

What about the uncertainty?

```{r}
big.tile <- big.krig |> st_join(NA_points)
ggplot() +
  geom_sf(data=big.tile, aes(geometry=tile, fill=var1.var), col=NA) +
  geom_sf(data=NA_boundary, fill=NA) +
  geom_sf(data=bats, aes(size=mass), shape=21) +
  coord_sf(crs = crs_for_map) +
  scale_fill_distiller(palette = "YlOrBr", limits=c(0.5,4))
```

By the looks our uncertainty has reduced where we're interpolating (i.e. where we have data, or between where we hvae data), but not by all that much. This makes sense as the model hasn't really improved the R^2 all that much! It has also made things worse where we are extrapolating - doing this with a linear model is a generally bad thing to do, as we're assuming that the linearity of the relationship holds outside of the data we observe, which might not be the case.
